PyPy's Approach to Virtual Machine Construction

### 2006-05-31 talk/dls2006/paper.tex
Added the screenshot.  Moved the figures around until they are reasonably placed.

\begin{figure}
\centering
\includegraphics[scale=0.667]{image/arch.pdf}
\caption{Overview of the translation process.}
\label{fig_arch}
\end{figure}

@@

\begin{figure}
\begin{verbatim}
def ll_append(lst, newitem):
  # Append an item to the end of the vector.
  index = lst.length       # get the 'length' field
  ll_resize(lst, index+1)  # call another helper
  itemsarray = lst.items   # get the 'items' field
  itemsarray[index] = item # behaves like a C array
\end{verbatim}
\caption{A helper to implement \texttt{list.append()}.}
\label{fig_llappend}
\end{figure}

@@

Note that such system code cannot typically be expressed as normal
RPython functions, because it corresponds to primitive operations at
that level.  As an aside, let us remark that the number of primitive
operations at RPython level is, comparatively speaking, quite large:
all list and dictionary operations, instance and class attribute
accesses, many string processing methods, a good subset of all Python
built-in functions...  Compared to other approaches [e.g.\ Squeak], we
do not try to minimize the number of primitives -- at least not at the
source level.  It is fine to have many primitives at any high enough
level, because they can all be implemented at the next lower level in
a way that makes sense to that level.  The key reason why this is not
burdensome is that the lower level implementations are also written in
Python -- with the only difference that they use (and have to be
typeable in) the lower-level type system.\footnote{This is not strictly
true: the type systems are even allowed to co-exist in the same
function.  The operations involving higher-level type systems are
turned into lower-level operations by the previous transformations in
the chain, which leave the already-low-level operations untouched.}

@@

The various analyses used -- from type inference to lifetime analysis -
are generally formulated as [abstract interpretation].  While this
approach is known to be less efficient than more tailored algorithms
like constraint-based type inference, we gain in freedom,
controllability and simplicity.  This proved essential in our overall
approach: as described in section \ref{systemprog}, we need to perform
type inference with many different type systems, the details of which
have evolved over time.

@@

We found in-process debugging sufficient and did not implement dumping
of any intermediate step to disk.  Figure 3 shows the control flow graph
obtained for a simple function -- this is a screenshot from our graph
viewer, used for debugging; basic block placement is performed by
Graphviz \cite{graphviz}.

\begin{verbatim}
::

    [figure 3: insert a nice pygame screenshot]
\end{verbatim}

@@

\begin{figure}
  \centering
  \caption{the interpreter and object spaces}
  \label{interpobjspace}
  \begin{tabular}{|c|c|} \hline
\multicolumn{2}{|c|}{forest of bytecode objects from the application}    \\ \hline
\multicolumn{2}{|c|}{Python bytecode interpreter}            \\ \hline
    Standard Object Space     &  Flow Object Space  \\ \hline
  \end{tabular}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[scale=0.75]{image/flowlattice.pdf}
\caption{The lattice order of the flow object space}
\label{flowlattice}
\end{figure*}

@@

@@

\begin{figure*}
\centering
\includegraphics[scale=0.75]{image/lattice1.pdf}
\caption{The overall lattice}
\label{latticeoverall}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[scale=0.75]{image/lattice2.pdf}
\caption{The part about instances and nullable instances, assuming a
  simple class hierarchy with only two direct subclasses of
  \texttt{object}.}
\label{latticedetail}
\end{figure*}

@@

\begin{table*}
% the layout of this table isn't wonderful, but it's probably OK.
\centering
\caption{Summary of interpreter performance}
\label{perfsumm}
\begin{tabular}{|l|c|c|} \hline
\textbf{Interpreter} & 
\textbf{Richards, Time/iteration} &
\textbf{Pystone, Iterations/second} \\ \hline
CPython 2.4.3                &   789ms  (1.0x) & 40322  (1.0x) \\ \hline
pypy-c                       &  4269ms  (5.4x) &  7587  (5.3x) \\ \hline
pypy-c-thread                &  4552ms  (5.8x) &  7122  (5.7x) \\ \hline
pypy-c-stackless             &  5121ms  (6.5x) &  6060  (6.7x) \\ \hline
pypy-c-gcframework           &  6327ms  (8.0x) &  4960  (8.1x) \\ \hline
pypy-c-stackless-gcframework &  8743ms (11.0x) &  3755 (10.7x) \\ \hline
pypy-llvm-c                  &  3797ms  (4.8x) &  7763  (5.2x) \\ \hline
pypy-llvm-c-prof             &  2772ms  (3.5x) & 10245  (3.9x) \\ \hline
\end{tabular}
\end{table*}

@@

\begin{table*}
\centering
\caption{Dividing overall translation time by stage}
\label{translationtimes}
\begin{tabular}{|p{0.65\linewidth}|c|} \hline
\textbf{Step} & \textbf{Time (minutes:seconds)} \\ \hline
Front-end                                 
(flow graphs and type inference)          
& 9:01 \\ \hline
LLTyper                                   
(from RPython-level to C-level graphs     
and data)                                 
& 10:38 \\ \hline
Various low-level optimizations           
(convert some heap allocations to local   
variables, inlining, ...)                 
& 6:51 \\ \hline
Database building                         
(this initial back-end step follows all   
graphs and prebuilt data structures       
recursively, assigns names, and orders    
them suitably for code generation)        
& 8:39 \\ \hline
Generating C source                       
& 2:25 \\ \hline
Compiling (\texttt{gcc -O2})              
& 3:23 \\ \hline
\end{tabular}
\end{table*}

>>>>

\begin{figure}
\centering
\includegraphics[scale=0.667]{image/arch.pdf}
\caption{overview of the translation process.}
\label{fig_arch}
\end{figure}

@@

\begin{figure}
\begin{verbatim}
def ll_append(lst, newitem):
  # Append an item to the end of the vector.
  index = lst.length       # get the 'length' field
  ll_resize(lst, index+1)  # call another helper
  itemsarray = lst.items   # get the 'items' field
  itemsarray[index] = item # behaves like a C array
\end{verbatim}
\caption{a helper to implement \texttt{list.append()}.}
\label{fig_llappend}
\end{figure}

@@

Note that such system code cannot typically be expressed as normal
RPython functions, because it corresponds to primitive operations at
that level.  As an aside, let us remark that the number of primitive
operations at RPython level is, comparatively speaking, quite large:
all list and dictionary operations, instance and class attribute
accesses, many string processing methods, a good subset of all Python
built-in functions...  Compared to other approaches
(e.g.\ Squeak \cite{Squeak}), we
do not try to minimize the number of primitives -- at least not at the
source level.  It is fine to have many primitives at any high enough
level, because they can all be implemented at the next lower level in
a way that makes sense to that level.  The key reason why this is not
burdensome is that the lower level implementations are also written in
Python -- with the only difference that they use (and have to be
typeable in) the lower-level type system.\footnote{This is not strictly
true: the type systems are even allowed to co-exist in the same
function.  The operations involving higher-level type systems are
turned into lower-level operations by the previous transformations in
the chain, which leave the already-low-level operations untouched.}

@@

The various analyses used -- from type inference to lifetime analysis -
are generally formulated as \textit{abstract interpretation.}  While this
approach is known to be less efficient than more tailored algorithms
like constraint-based type inference \cite{Constraint}, we gain in freedom,
controllability and simplicity.  This proved essential in our overall
approach: as described in section \ref{systemprog}, we need to perform
type inference with many different type systems, the details of which
have evolved over time.

@@

We found in-process debugging sufficient and did not implement dumping
of any intermediate step to disk.  Figure \ref{fig_screenshot}
shows the control flow graph
obtained for a simple function -- this is a screenshot from our graph
viewer, used for debugging; basic block placement is performed by
Graphviz \cite{graphviz}.

\begin{figure}
\includegraphics[scale=0.59]{image/screenshot-bw.pdf}
\caption{the control flow graph of a simple function, computed and
displayed by our tool-chain.}
\label{fig_screenshot}
\end{figure}

@@

\begin{figure}
  \centering
  \label{interpobjspace}
  \begin{tabular}{|c|c|} \hline
\multicolumn{2}{|c|}{\vphantom{\Large X}forest of bytecode objects from the application}    \\ \hline
\multicolumn{2}{|c|}{\vphantom{\Large X}Python bytecode interpreter}            \\ \hline
    \vphantom{\Large X} Standard Object Space     &  Flow Object Space  \\ \hline
  \end{tabular}
  \caption{the interpreter and object spaces.}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[scale=0.75]{image/flowlattice.pdf}
\caption{the lattice order of the flow object space.}
\label{flowlattice}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[scale=0.667]{image/lattice1.pdf}
\caption{the overall lattice.}
\label{latticeoverall}
\end{figure*}

@@

\begin{figure}
\centering
\includegraphics[scale=0.667]{image/lattice2.pdf}
\caption{the part about instances and nullable instances, assuming a
  simple class hierarchy with only two direct subclasses of
  \texttt{object}.}
\label{latticedetail}
\end{figure}

@@

\begin{table*}
% the layout of this table isn't wonderful, but it's probably OK.
\centering
\label{perfsumm}
\begin{tabular}{|l|c|c|} \hline
\textbf{Interpreter} & 
\textbf{Richards, Time/iteration} &
\textbf{Pystone, Iterations/second} \\ \hline
CPython 2.4.3                &   789ms  (1.0x) & 40322  (1.0x) \\ \hline
pypy-c                       &  4269ms  (5.4x) &  7587  (5.3x) \\ \hline
pypy-c-thread                &  4552ms  (5.8x) &  7122  (5.7x) \\ \hline
pypy-c-stackless             &  5121ms  (6.5x) &  6060  (6.7x) \\ \hline
pypy-c-gcframework           &  6327ms  (8.0x) &  4960  (8.1x) \\ \hline
pypy-c-stackless-gcframework &  8743ms (11.0x) &  3755 (10.7x) \\ \hline
pypy-llvm-c                  &  3797ms  (4.8x) &  7763  (5.2x) \\ \hline
pypy-llvm-c-prof             &  2772ms  (3.5x) & 10245  (3.9x) \\ \hline
\end{tabular}
\caption{summary of interpreter performance.}
\end{table*}

@@

\begin{table*}
\centering
\label{translationtimes}
\begin{tabular}{|p{0.65\linewidth}|c|} \hline
\textbf{Step} & \textbf{Time (minutes:seconds)} \\ \hline
Front-end                                 
(flow graphs and type inference)          
& 9:01 \\ \hline
LLTyper                                   
(from RPython-level to C-level graphs     
and data)                                 
& 10:38 \\ \hline
Various low-level optimizations           
(convert some heap allocations to local   
variables, inlining, ...)                 
& 6:51 \\ \hline
Database building                         
(this initial back-end step follows all   
graphs and prebuilt data structures       
recursively, assigns names, and orders    
them suitably for code generation)        
& 8:39 \\ \hline
Generating C source                       
& 2:25 \\ \hline
Compiling (\texttt{gcc -O2})              
& 3:23 \\ \hline
\end{tabular}
\caption{dividing overall translation time by stage.}
\end{table*}
